Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-09-06 analysis/paper/paper.Rmd
Corrected reference key and added code for performing pca plots in R

```{r load-packages-and-data}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(ggsci)

@@

@@

#Filter data to focus on quartzite points, point fragments and preforms made from quartz/quartzite materials
Points <-
  XRF.csv %>%
  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

#Import loadings from Evince
loadings <-
 as.data.frame(read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data/XRF/XRF_P_values.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
```
<!-- The actual document text starts here: -->

@@

```{r distribution-by-parish, fig.cap= paste("Distribution of the", length(sample_uniq), "points and preforms across the parishes in Västerbotten, Sweden. Note that while Lycksele feature a large number of samples", length(lycksele_301), "of these are from the site of Lycksele 301"), fig.width=12, fig.height=10, fig.fullwidth=TRUE}
#Stacked bar chart with counts of samples per parish and type
ggplot(Points2, aes(x= fct_rev(parish), fill = type)) +
  geom_bar() +
  theme_minimal() +
  labs(y = "Count",
       fill = "Type") +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(face = "bold"),
    axis.title.y=element_blank(),
    axis.text.y = element_text(size = 10, face = "bold", colour = "black"),
    axis.title.x = element_text(size = 12, face = "bold", colour = "black")) +
  scale_fill_jco() +
  coord_flip() +
  geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)

@@

```{r nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum. Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant. A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
```

@@

```{r raman-average, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_4_raman_spectra_dark.jpg')
```

@@

```{r raman-pca1, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components of the raman analysis. The materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G band peaks around 1300 cm-1 and 1600 cm-1. Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_6_raman_average_pca_p1_p2.jpg')
```

```{r raman-pca2, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 3rd components of the raman analysis. The materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G band peaks around 1300 cm-1 and 1600 cm-1. Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_7_raman_average_pca_p1_p3.jpg')
```

@@

```{r xrf-violin-plot, fig.width=15, fig.height=10, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
#Prepare data in long format for violin plots
Points_long <-
  Points %>%
  select(X12Mg,X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti) %>%
  gather(key = "Element", value = "Value")

@@

#Lookup for element labeller
Elements <- c(
  '12Mg' = "Mg",
  '13Al' = "Al",
  '14Si' = "Si",
  '15P' = "P",
  '16S'= "S",
  '19K' = "K",
  '20Ca' ="Ca",
  '23V' = "V",
  '25Mn' = "Mn",
  '26Fe' = "Fe",
  '30Zn' = "Zn",
  '38Sr' = "Sr",
  '39Y' = "Y",
  '40Zr' = "Zr",
  '56Ba' = "Ba",
  '81Ti' = "Ti"
  )

@@

```{r xrf-loadings, fig.width=8, fig.height=8, fig.fullwidth=TRUE, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
#Prepare plots for 3-plot bargraph figure of loadings
#PC1 loadings
P1 <-
  ggplot(loadings, aes(x = fct_rev(Element), y = P1)) +
    geom_bar(
      stat = "identity", position = "identity",
      color = "black", fill = "darkgrey") +
  theme_minimal() +
  theme(plot.margin = margin(10,10,10,10),
        axis.title.y=element_blank(),
        axis.text.y = element_text(size=14, face="bold", colour = "black"),
        axis.title.x = element_text(size = 14, face = "bold", colour = "black")) +
  coord_flip()
#PC2 loadings
P2 <-
  ggplot(loadings, aes(x = fct_rev(Element), y = P2)) +
  geom_bar(
    stat = "identity", position = "identity",
    color = "black", fill = "darkgrey") +
  theme_minimal() +
  theme(plot.margin = margin(10,10,10,10),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.title.x = element_text(size = 14, face = "bold", colour = "black")) +
  coord_flip()
#PC3 loadings
P3 <-
  ggplot(loadings, aes(x = fct_rev(Element), y = P3)) +
  geom_bar(
    stat = "identity", position = "identity",
    color = "black", fill = "darkgrey") +
  theme_minimal() +
  theme(plot.margin = margin(10,10,10,10),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14, face = "bold", colour = "black")) +
  coord_flip()

#Composite figure of above bar graphs
grid.arrange(P1, P2, P3, ncol=3)
```

>>>>

```{r load-packages-and-data}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(ggsci)
library(prospectr)
library(patchwork)
library(factoextra)
library(ggbiplot)

@@

#Import nir data, set empty fields to NA
nir.csv <-
  read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/derived_data//NIR_master_averaged_20210827.csv", sep = ";",
            dec = ",", header = TRUE, na = c("","NA","NULL",NULL))

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/derived_data//raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, na = c("","NA","NULL",NULL))

@@

#Filter xrf, nir and raman data to focus on points and preforms made from quartz/quartzite material 
Points.xrf <-
  XRF.csv %>%
  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

Points.nir <-
  nir.csv %>%
  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

Points.raman <-
  raman.csv %>%
  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
         type == "Point" | type == "Point fragment" | type == "Preform")

#Fill the NA fields in the munsell_hue column and mark them as translucent samples
Points.nir[8][is.na(Points.nir[8])] <- "Translucent"
Points.raman[8][is.na(Points.raman[8])] <- "Translucent"
Points.xrf[8][is.na(Points.xrf[8])] <- "Translucent"

#prepare color/fill and symbols for score plots
colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
shapes <- c(24,25,23,22,21)

#Import loadings from Evince
loadings <-
 as.data.frame(read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data/XRF/XRF_P_values_long.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
```
<!-- The actual document text starts here: -->

@@

```{r distribution-by-parish, fig.cap= paste("Distribution of the", length(sample_uniq), "points and preforms across the parishes in Västerbotten, Sweden. Note that while Lycksele feature a large number of samples", length(lycksele_301), "of these are from the site of Lycksele 301"), fig.width=12, fig.height=10, fig.fullwidth=TRUE}

#Stacked bar chart with counts of samples per parish and type
ggplot(Points2, aes(x= fct_rev(parish), fill = type)) +
  geom_bar() +
  theme_minimal() +
  labs(y = "Count",
       fill = "Type") +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(face = "bold"),
    axis.title.y=element_blank(),
    axis.text.y = element_text(size = 10, face = "bold", colour = "black"),
    axis.title.x = element_text(size = 12, face = "bold", colour = "black")) +
  scale_fill_jco() +
  coord_flip() +
  geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)

@@

<!--```{r nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the #spectrum. Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant. A trend #of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
#knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
``` -->
```{r nir-pca, fig.width=20, fig.height=8, fig.cap= "PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum. Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant. A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}

#perform PCA with SNV normalization and mean-center
nir.pca <-
  prcomp(Points.nir[,c(678:2178)], center = TRUE, scale = FALSE)

#prepare a basic score plot with fviz_pca_ind using PC1 and PC2
basic_plot1 <-
  fviz_pca_ind(nir.pca, axes = c(1,2), label="none")

#prepare labels for PCs
pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=1)
pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=1)
pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 1 000 - 2 500 nm") +
  labs(x = pc1lab,
       y = pc2lab) +
  scale_shape_manual(name = "Material",
                     values=shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
        legend.title = element_text(size = 12, face = "bold", colour = "black"),
        legend.background = element_rect(linetype = "solid", color = "black"),
        legend.position = "bottom")
```

@@

<!--```{r raman-average, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_4_raman_spectra_dark.jpg')
```-->
```{r raman-average, fig.width=20, fig.height=15, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}

#perform PCA with SNV normalization and mean-center
raman.pca <-
  prcomp(standardNormalVariate(X = Points.raman[,c(29:320)]), center = TRUE, scale = FALSE)

#prepare different basic score plots with fviz_pca_ind
basic_plot1 <-
  fviz_pca_ind(raman.pca, axes = c(1,2), label="none")

basic_plot2 <-
  fviz_pca_ind(raman.pca, axes = c(1,3), label="none")

#prepare labels for PCs
pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
pc3lab <- paste0("PC3 (",as.character(pc3var),"%)")

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
p1 <- ggplot(cbind(basic_plot1$data, Points.raman[, c(8,10)]),
       aes(x=x, y=y, 
           shape = material, 
           fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Raman") +
  labs(x = pc1lab,
       y = pc2lab) +
  scale_shape_manual(name = "Material",
                     values=shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(legend.position = "none")

#bind the basic fviz plot for PC 1 and 3, and use as basis for a more customizeable plot in ggpplot
p2 <- ggplot(cbind(basic_plot2$data, Points.raman[, c(8,10)]),
       aes(x=x, y=y, 
           shape = material, 
           fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  labs(x = pc1lab,
       y = pc3lab) +
  scale_shape_manual(name = "Material",
                     values=shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(legend.position = "none")

combined <-
  p1 + p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
                  axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
                  axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
                  legend.title = element_text(size = 12, face = "bold", colour = "black"),
                  legend.background = element_rect(linetype = "solid", color = "black"),
                  legend.position = "bottom")

combined +
  plot_layout(nrow = 2, guides = "collect")
```

@@

```{r raman-pca1, fig.fullheight=TRUE, fig.fullwidth=TRUE, fig.cap="PCA score plots for the raman analysis. Along the 1st component (P1) materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G bands around 1300 cm-1 and 1600 cm-1. Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak A) Plot of 1st and 2nd components; noisier spectra drift outwards along P2 B) Plot of 1st and 3rd components; noisier spectra are separated out from other samples, but spectra with peaks around 700-800 cm-1 are drawn towards the bottom right."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_6_raman_average_pca_p1_p2_p3.jpg')
```

@@

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
#Prepare data in long format for violin plots
Points_long <-
  Points.xrf %>%
  select(X12Mg,X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti) %>%
  gather(key = "Element", value = "Value")

@@

#Lookup for element labeller
#Ändra ordning så att de mer avvikande (Si, Mn etc) hamnar uppe till vänster
Elements <- c(
  '12Mg' = "Mg",
  '13Al' = "Al",
  '14Si' = "Si",
  '15P' = "P",
  '16S'= "S",
  '19K' = "K",
  '20Ca' ="Ca",
  '23V' = "V",
  '25Mn' = "Mn",
  '26Fe' = "Fe",
  '30Zn' = "Zn",
  '38Sr' = "Sr",
  '39Y' = "Y",
  '40Zr' = "Zr",
  '56Ba' = "Ba",
  '81Ti' = "Ti"
  )

@@

```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
#bar graph of the loadings (P1-P3) for the XRF data  
loadings %>%
  ggplot(aes(x=fct_rev(Element), y=Value)) +
    geom_bar(
      stat = "identity", position = "identity",
      color = "black", fill = "darkgrey",
      width = 0.5) +
  facet_wrap( ~ P, scales = "free") +
  theme_bw() +
  theme(plot.margin = margin(10,10,10,10),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10, face="bold", colour = "black"),
        axis.title.x = element_blank(),
        strip.text.x = element_text(size=14, face="bold", colour="black")) +
  coord_flip()
